Gap formation and stability in non-isothermal protoplanetary discs

### 2014-12-14 paper/model.tex
final edit by MKL

\subsection{Parametrized cooling}
In this work we only vary one control parameter: the cooling
time. %The disc parameters are $Q_o$, which characterizes the strength of self-gravity; 
%and $\beta$ which sets the cooling time. Two levels of self-gravity is considered: 
%a light disc with $Q_o=8$ and a heavy disc with $Q_o=1.5$. 
The cooling parameter $\beta$ is chosen indirectly  through the parameter
$\tilde{\beta}$ such that

@@

\subsection{Diagonistic measures}

\subsubsection{Generalized potential vorticity}

The generalized potential vorticity is defined as
\begin{align}
  \tilde{\eta} = \frac{\kappa^2}{2\Omega\Sigma}\times S^{-2/\gamma}, 
\end{align}
where $\kappa^2 = r^{-3}\partial_r(r^4\Omega^2)$ is the square of the
epicycle frequency, $\Omega=v_\phi/r$ is the angular speed, and
$S\equiv p/\Sigma^\gamma$ is the entropy. The first factor is the
usual potential vorticity (PV, or vortensity).

The generalized PV appears in the description of the linear stability
of radially-structured adiabatic discs \citep{lovelace99,li00}, where
the authors show an extremum in $\teta$ may lead to a dynamical
instability, the RWI. In a barotropic disc, the entropy factor is
absent and the important quantity is the PV.

\subsubsection{Fourier modes} 
The RWI is characterized by exponetially
growing perturbations. Though in this paper we do not consider a
formal linear instability calculation, modal analysis will be useful
to analyse the growth of perturbations with different azimuthal
wavenumbers, which is associated with the number of vortices initially
formed by the RWI.

The Fourier transform of the time-dependent surface density is
\begin{align}\label{fouriertransform}
  \Sigma_m(r,t) = \int_{0}^{2\pi}
  \Sigma(r,\phi,t) \, \mathrm{e}^{-\mathrm{i}m\phi} \, \mathrm{d}\phi 
\end{align} 
where $m$ is the aziumuthal wave number. We define the growth rate
$\gamma$ of the $m^\mathrm{th}$ component of the surface density
through 
% and $\sigma_m$ is the complex frequency
%of the $m \mathrm{th}$ mode. Since $\sigma_m$ is a value dependent on
%$r$, a more useful diagnostic is the mode growth rate defined as  
\begin{align}\label{growth}
  \frac{d \langle|\Sigma_m|\rangle_r }{dt}= \gamma \langle|\Sigma_m|\rangle_r 
\end{align}
where %$\gamma$ is the real part of the $\sigma_m$ and 
$\langle|\cdot|\rangle_r$ denotes the average of the absolute value
over a radial region of interest. By using Eq.~\ref{growth} the growth
rates of the unstable modes can be found from successive spatial
Fourier transforms over an appropriate period of time.

>>>>

\subsection{Cooling prescription}
In this work we only vary one control parameter: the cooling
time. %The disc parameters are $Q_o$, which characterizes the strength of self-gravity; 
%and $\beta$ which sets the cooling time. Two levels of self-gravity is considered: 
%a light disc with $Q_o=8$ and a heavy disc with $Q_o=1.5$. 
The cooling parameter $\beta$ is chosen indirectly  through the parameter
$\tilde{\beta}$ such that

@@

\subsection{Diagnostic measures}

\subsubsection{Generalised potential vorticity}

The generalised potential vorticity is defined as
\begin{align}
  \tilde{\eta} = \frac{\kappa^2}{2\Omega\Sigma}\times S^{-2/\gamma}, 
\end{align}
where $\kappa^2 = r^{-3}\partial_r(r^4\Omega^2)$ is the square of the
epicyclic frequency, $\Omega=v_\phi/r$ is the angular speed, and
$S\equiv p/\Sigma^\gamma$ is the entropy. The first factor is the
usual potential vorticity (PV, or vortensity).

The generalised PV appears in the description of the linear stability
of radially-structured adiabatic discs \citep{lovelace99,li00}, where
the authors show an extremum in $\teta$ may lead to a dynamical
instability, the RWI. In a barotropic disc, the entropy factor is
absent and the important quantity is the PV.

\subsubsection{Fourier modes} 
The RWI is characterised by exponentially
growing perturbations. Though in this paper we do not consider a
formal linear instability calculation, modal analysis will be useful
to analyse the growth of perturbations with different azimuthal
wavenumbers, which is associated with the number of vortices initially
formed by the RWI.

The Fourier transform of the time-dependent surface density is
\begin{align}\label{fouriertransform}
  \Sigma_m(r,t) = \int_{0}^{2\pi}
  \Sigma(r,\phi,t) \, \mathrm{e}^{-\mathrm{i}m\phi} \, \mathrm{d}\phi 
\end{align} 
where $m$ is the azimuthal wave number. We define the growth rate
$\gamma$ of the $m^\mathrm{th}$ component of the surface density
through 
% and $\sigma_m$ is the complex frequency
%of the $m \mathrm{th}$ mode. Since $\sigma_m$ is a value dependent on
%$r$, a more useful diagnostic is the mode growth rate defined as  
\begin{align}\label{growth}
  \frac{d \langle|\Sigma_m|\rangle_r }{dt}= \gamma \langle|\Sigma_m|\rangle_r 
\end{align}
where %$\gamma$ is the real part of the $\sigma_m$ and 
$\langle|\cdot|\rangle_r$ denotes the average of the absolute value
over a radial region of interest. By using Eq.~\ref{growth} the growth
rates of the unstable modes can be found from successive spatial
Fourier transforms over an appropriate period of time.

### 2014-12-14 paper/paper.tex
Update
final edit by MKL

@@

\maketitle
\begin{abstract}
  Several observations of transition discs show lopsided
  dust-distributions. A potential explanation is the formation of a
  large-scale vortex acting as a dust-trap at the edge of a gap opened
  by a giant planet. Numerical models of gap-edge vortices have
  thus far employed locally isothermal discs, but the 
  theory of this vortex-forming or `Rossby wave' instability was
  originally developed for adiabatic discs.  
  We generalise the study of planetary gap stability to non-isothermal
  discs using customised numerical simulations of disc-planet
  systems where the planet opens an unstable gap. 
  We include in the energy equation a simple cooling function with
  cooling timescale $t_c=\beta\Omega_k^{-1}$, where $\Omega_k$ is
  the Keplerian frequency, and examine the effect of $\beta$ on the
  stability of gap edges and vortex lifetimes. We find increasing
  $\beta$ lowers the growth rate of non-axisymmetric perturbations, and the
  dominant azimuthal wavenumber $m$ decreases.  
  We find a quasi-steady state consisting of one  
  large-scale, over-dense vortex circulating the outer gap edge, typically  
  lasting $O(10^3)$ orbits. 
  Vortex lifetimes were found to increase with cooling times up to an
  optimal value, beyond which vortex lifetimes decrease. This
  non-monotonic dependence is qualitatively consistent with 
  recent studies using strictly isothermal discs that vary the disc aspect ratio. 
%  presented in Fu et al. (2014, ApJL, 788, L41). 
\end{abstract}

@@

\section{Introduction}\label{intro}
%{
%\bf 
%observational motivation: asymmetric dust distributions. 
%theoretical motivations: generalize gap stability to include energy equation. 
%}
The interaction between planets and protoplantary discs plays an
important role in the theory of planet formation and disc 
evolution. Disc-planet interaction may lead to the orbital migration
of protoplanets and modify the structure of
protoplanetary discs  \citep[see][for a recent review]{baruteau13}.  
%\citep[][see \cite{baruteau13} for a recent
%  review]{goldreich79, lin79, goldreich80,lin86}.

@@

This paper is organised as follows. In \S\ref{model} we describe the
equations governing the disc-planet system and initial conditions. Our
numerical approach, including diagnositic measures, are given in
\S\ref{method}. We present results from two sets of numerical
experiments. In \S\ref{linear1} we use disc-planet interaction to set
up discs with gaps, but study their stability without further
influence from the planet. %stability through background state 
We then perform long-term disc-planet simulations to examine the
lifetime of gap-edge vortices in \S\ref{nonlinear},  
as a function of the imposed cooling rate. We conclude and summarize
in \S\ref{summary} with a discussion of important caveats.

@@

%\section*{Acknowledgments}

>>>>

\def\la{\left\langle\rule{0pt}{3em}}
\def\ra{\right\rangle}

@@

\maketitle
\begin{abstract}
  Several observations of transition discs show lopsided
  dust-distributions. A potential explanation is the formation of a
  large-scale vortex acting as a dust-trap at the edge of a gap opened
  by a giant planet. Numerical models of gap-edge vortices have
  thus far employed locally isothermal discs, but the 
  theory of this vortex-forming or `Rossby wave' instability was
  originally developed for adiabatic discs.  
  We generalise the study of planetary gap stability to non-isothermal
  discs using customised numerical simulations of disc-planet
  systems where the planet opens an unstable gap. 
  We include in the energy equation a simple cooling function with
  cooling timescale $t_c=\beta\Omega_k^{-1}$, where $\Omega_k$ is
  the Keplerian frequency, and examine the effect of $\beta$ on the
  stability of gap edges and vortex lifetimes. We find increasing
  $\beta$ lowers the growth rate of non-axisymmetric perturbations, and the
  dominant azimuthal wavenumber $m$ decreases.  
  We find a quasi-steady state consisting of one  
  large-scale, over-dense vortex circulating the outer gap edge, typically  
  lasting $O(10^3)$ orbits. 
  Vortex lifetimes were found to generally increase with cooling times up to an
  optimal value, beyond which vortex lifetimes decrease. This
  non-monotonic dependence is qualitatively consistent with 
  recent studies using strictly isothermal discs that vary the disc aspect ratio. 
%  presented in Fu et al. (2014, ApJL, 788, L41). 
\end{abstract}

@@

\section{Introduction}\label{intro}
%{
%\bf 
%observational motivation: asymmetric dust distributions. 
%theoretical motivations: generalize gap stability to include energy equation. 
%}
The interaction between planets and protoplanetary discs plays an
important role in the theory of planet formation and disc 
evolution. Disc-planet interaction may lead to the orbital migration
of protoplanets and modify the structure of
protoplanetary discs  \citep[see][for a recent review]{baruteau13}.  
%\citep[][see \cite{baruteau13} for a recent
%  review]{goldreich79, lin79, goldreich80,lin86}.

@@

This paper is organised as follows. In \S\ref{model} we describe the
equations governing the disc-planet system and initial conditions. Our
numerical approach, including diagnostic measures, are given in
\S\ref{method}. We present results from two sets of numerical
experiments. In \S\ref{linear1} we use disc-planet interaction to set
up discs with gaps, but study their stability without further
influence from the planet. %stability through background state 
We then perform long-term disc-planet simulations to examine the
lifetime of gap-edge vortices in \S\ref{nonlinear},  
as a function of the imposed cooling rate. We conclude and summarise
in \S\ref{summary} with a discussion of important caveats.

@@

\section*{Acknowledgments}
This project was initiated at the Canadian Institute for Theoretical
Astrophysics (CITA) 2014 summer student programme.  
Computations were performed on the GPC supercomputer at the
SciNet HPC Consortium. SciNet is funded by: the Canada Foundation for
Innovation under the auspices of Compute Canada; the Government of
Ontario; Ontario Research Fund - Research Excellence; and the
University of Toronto.
